Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-26 newpaper/paper.tex
first draft of paper version

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Applicative}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}
\newcommand{\foldr}{\msf{foldr}}
\newcommand{\foldl}{\msf{foldl}}
\newcommand{\rep}{\msf{rep}}
\newcommand{\concatMap}{\msf{concatMap}}

@@

\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{{<\!\!*\!\!>}}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\MonadState}{\msf{MonadState}}
\newcommand{\MonadReader}{\msf{MonadReader}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\Either}{\msf{Either}}
\newcommand{\Left}{\msf{Left}}
\newcommand{\Right}{\msf{Right}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}

@@

GHC's optimization efforts are respectable, and it has gotten quite good over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation, but without any strong guarantee, and their
advanced usage requires knowledge of GHC internals. For example, correctly
specifying the \emph{ordering} of certain rule applications is often needed. We
also have to care about formal function arities. Infamously, the function
composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to \vf\,(\vg\,\vx)$ in the
base libraries, instead of as $(.)\,\vf\,\vg\,\vx = \vf\,(\vg\,\vx)$, to get better inlining
behavior \cite{TODO}. It remains a common practice in high-performance Haskell
programming to manually review GHC's optimized code output.

@@

The idea of \textbf{two-level type theory} (2LTT) is to use a highly expressive
dependent type theory for compile-time computation, but generate code in
possibly different, simpler object language. In this paper, we use 2LTT to sweeten the
deal of staged compilation, aiming for a combination of strong guarantees, good
ergonomics, high level of abstraction and easy-to-optimize code output.

We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists
of
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      in the downstream pipeline, but it lacks many convenience features.
\item A standard Martin-Löf type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}

Since the object language is first-order, we guarantee that all programs in CFTT
can be ultimately compiled without any dynamic closures, using only calls and
jumps to statically known code. Why focus on closures?  They are the
foundation to almost all abstraction tools in functional programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods.
\item Functors and first-class modules in OCaml and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

@@

%% \todo{Essential closures: CPS? Threaded interpreters?}

%% Whole-program defunctionalization is also possible, and it is notably used by
%% the MLton compiler \cite{TODO}. While this can be practical and effective, it
%% can be also expensive and require whole-program processing. Also, conceptually
%% speaking, it does not eliminate closures but instead makes them more transparent
%% to optimizations. In this paper we do not use defunctionalization.

@@

\begin{itemize}
\item In Section \ref{TODO} we present the two-level type theory CFTT, where the
  object language is first-order simply-typed and the metalanguage is
  dependently typed. The object language admits an operational semantics without
  runtime closures, and can be compiled in a way such that all function calls
  are statically known. We provide a supplementary Agda formalization of the
  operational semantics.
\item In Section \ref{TODO} we build a monad transformer library. We believe
  that this is a good demonstration, because monads and monad transformers are
  the most widely used effect system in Haskell, and at the same time their
  compilation to efficient code can be surprisingly difficult. The main idea is
  to have monads and monad transformers \emph{only} in the metalanguage, and try
  to express as much as possible at the meta level, and funnel the object-meta
  interactions through specific \emph{binding-time improvements}.  We show how
  almost all definitions from the Haskell ecosystem of monad transformers can be
  reused in the meta level of CFTT. We also develop let-insertion, join points
  and case switching on object-level data, in the generality of monad transformers.
\item In Section \ref{TODO} we build a pull-based stream fusion library. Here,
  the motivation is to demonstrate essential usage of dependent types to
  concisely solve a difficult problem, namely guaranteed fusion for unrestricted
  combinations of $\concatMap$ and $\mit{zip}$. We use a state machine
  representation that is based on \emph{sums-of-products} of object-level
  values. We show that CFTT is compatible with a \emph{generativity} axiom,
  which internalizes the fact that metaprograms cannot inspect the structure of
  object-level terms. We use this to show that the universe of sums-of-products
  is closed under $\Sigma$-types. This in turn enables a very
  concise definition of $\concatMap$.
\item
  We adapt the contents of the paper to a Template Haskell, with some
  modifications, simplifications and fewer guarantees about generated code.  In
  particular, Haskell does not have enough dependent types for the simple
  $\concatMap$ definition, but we can still work around this limitation. We also
  provide a more faithful Agda embedding of CFTT and our libraries. Here, the
  object theory is embedded as a collection of postulated operations, and we can
  use Agda's normalization command to print out generated object code.
\end{itemize}

@@

$\bs{\MTy}$ is the universe of types in the compile-time language. We will often
use the term ``metatype'' to refer to inhabitants of $\MTy$, and use
``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
functions, products and indexed inductive types \cite{TODO}.

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
relevant to the current paper, so we shall omit levels. Note that universe
levels are orthogonal to staging; they are instead used to increase the strength
of the system while maintaining logical consistency.

@@

We use Haskell-like newtype notation, such as in $\newtype \msf{Wrap}\,\vA =
\msf{Wrap}\,\{\mit{unWrap} : \vA \}$, and also a similar notation for (dependent)
record types, for instance as in
\[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]

All construction and elimination rules for type formers in $\MTy$ stay within
$\MTy$. For example, induction on meta-level values can only produce meta-level
values.

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so so we have $\Ty : \MTy$. Similarly as in the case of $\MTy$, all
construction and elimination rules of the object language stay within $\Ty$.
However, we further split $\Ty$ to two sub-universes.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
to define computations and values as well, and can be used to shadow binders:
\begin{alignat*}{3}
  &\vf : \Nat \to \Nat\\
  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
\end{alignat*}
We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
assumed to be erased at runtime. In the Haskell implementation they are
important for guiding type class resolution, and we think that the explicit
wrapping makes many definitions more comprehensible in CFTT as well.

@@

However, the call-by-name semantics lets us transform $\vf$ to $\lam
\vb\,\vx.\,\case \vb\,\of \True \to \vx + 10; \False \to \vx * 10$, and more
generally we can transform programs so that every function call is
\emph{saturated}. This means that every function call is of the form
$\vf\,\vt_1\,\vt_2\,...\,\vt_n$, where $\vf$ is a function variable and the
definition of $\vf$ immediately $\lambda$-binds $n$ arguments. We do not detail
this here. We provide formal syntax and operational semantics of the object
language in the Agda supplement. We formalized the specific translation steps
that are involved in call saturation, but only specified the full translation
informally.

@@

\subsubsection{Comparison to call-by-push-value}
We took inspiration from call-by-push-value (CBPV), and there are similarities,
but there are also significant differences. Both systems have a
value-computation distinction, with call-by-name computations and call-by-value
values. However, our object theory supports variable binding at arbitrary types
while CBPV only supports value variables. In CBPV, a let-definition for a
function is only possible by first packing it up as a closure value (or
``thunk''), which clearly does not suit us. In future work we could make
a closer look at the connections to CBPV.

%% \subsubsection{Spectrum of possible object languages}
%% There is a trade-off between making the object language more restricted, and
%% thus easier to compile, and making metaprogramming more convenient. We will see
%% that the ability to insert $\mbf{let}$-s without restriction is very convenient,
%% and likewise the ability to have $\lambda$-abstraction under $\mbf{case}$
%% bodies, although these features necessitate more downstream processing, to bring
%% programs to saturated form. In this paper we go for metaprogramming convenience,

@@

\begin{itemize}
\item For $\vA : \Ty$, we have $\Lift \vA : \MTy$, pronounced as ``lift $\vA$''. This is
      the type of metaprograms that produce $\vA$-typed object programs.
\item For $\vA : \Ty$ and $\vt : \vA$, we have $\ql \vt \qr : \Lift \vA$, pronounced ``quote $\vt$''. This
      is the metaprogram which immediately returns $\vt$.
\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice $\vt$''. This
  inserts the result of a metaprogram into an object term. \emph{Notation:}
  splicing binds stronger than function application, so $\vf\,\spl \vx$ is parsed as
  $\vf\,(\spl \vx)$.
\item We have $\ql \spl \vt \qr \equiv \vt$ and $\spl \ql \vt \qr \equiv \vt$ as definitional equalities.
\end{itemize}

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. \textbf{Staging} means running all metaprograms in splices and
inserting their output into object code, keeping all object-level top entries
and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
and produces output which is purely in the object-level fragment, with no
metatypes and metaprograms remaining. The output is guaranteed to be
well-typed. This staging is formalized in detail in \cite{TODO}. In Section
\ref{TODO} we describe the modifications to ibid.\ that are used in this paper.

@@

\subsection{Semantics of Staging}

We give a short summary of the semantics of staging here. We reuse most of the
semantics from \cite{TODO}, also borrowing notation and definitions, and only
mention additions and differences. The reader may refer to ibid.\ for technical
details.

Staging is given by evaluation of CFTT into a presheaf model, where the base
category is given as follows: objects are typing contexts of the object language
and morphisms are parallel substitutions. The equality of morphisms is given by
strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
which is constantly the set of object-theoretic types. Soundness of staging is
proven by a separate logical relations model, internally to the mentioned
presheaf model. Soundness means that the output of staging, viewed as a CFTT
term, is definitionally equal to the input.

The main addition in CFTT is the assumption of all inductive families in the
meta-level. However, based on the results in \cite{TODO}, it is enough to
extend the meta-level syntax in \cite{TODO} with W-types and identity types to
derive all inductive families, so we consider these in the following.

\subsubsection{W-types} W-types have a standard object-wise inductive definition in
the presheaf model \cite{TODO}. Additionally, we need to give a logical relation
interpretation to W-types, as required for the soundness proof for staging. In
Section 5.2.\ of \cite{TODO} a serialization map is used to define the relation
for natural numbers, however, there is no analogous map for W-types because we
cannot build finite terms from infinitely branching trees.

@@

\subsubsection{Identity type} Our treatment of the identity type is
influenced by the following: we would like to assume identities as axioms in
CFTT, which are true in the presheaf model (i.e.\ they are true at
staging-time), without blocking staging as a computation. In Section \ref{TODO}
we use such an axiom to good effect. However, axioms do block computation in
CFTT up to definitional equality, which implies that the soundness of staging
fails. Our solution:
\begin{itemize}
\item Have a standard intensional identity type in CFTT, perhaps with uniqueness of identity proofs assumed.
\item Before staging, translate CFTT to CFTT plus equality reflection. Operationally,
      this means erasing all identity proofs and transports.
\item Perform staging after erasure of identity proofs.
\end{itemize}
In the presheaf model, we interpret the extensional identity in a standard way,
as strict equality of sections of presheaves. With this, soundness of staging is
trivial to show; we can pick the trivial relation for the interpretation of
identity. Hence, we weaken the statement of soundness: every CFTT program is
convertible to the staging output, up to erasure of identity proofs.

@@

In this section we build a library for monads and monad transformers. We believe
that this is a good demonstration of CFTT-s abilities, since monads are
ubiquitous in Haskell programming, and they also introduce a great amount of
abstraction that should be optimized away.

@@

In the staged compilation and partial evaluation literature, the term
\emph{binding time improvement} is used to refer to such conversions, where
the ``improved'' version supports more compile-time computation. A general
strategy for generating efficient ``fused'' programs, is to try to work as much
as possible with improved representations, and only convert back to object code
at points where runtime dependencies are unavoidable. Let us look at
binding-time-improvement for product types now:
\begin{alignat*}{3}
  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
\end{alignat*}

%% We use a Haskell-style type class for binding time improvements:
%% \begin{alignat*}{3}
%%   &\hspace{-8em}\mathrlap{\tyclass\,\Improve\,(A : \Ty)\,(B : \MTy)\,\where} \\
%%   &\hspace{-8em}\ind \mup   &&: \Up A \to B\\
%%   &\hspace{-8em}\ind \mdown &&: B \to \Up A
%%   &\\
%%   &\hspace{-8em}\mathrlap{\instance \Improve\,(A \to B)\,(\Up A \to \Up B)\,\where\,\,...}
%% \end{alignat*}
%% We will use type classes in an informal way, without precisely specifying how
%% they work. However, in the Haskell adaptation of this paper we do make crucial
%% use of type classes. There are some necessary differences between the CFTT and
%% Haskell versions though, which we will summarize in Section \ref{TODO}. Let us
%% look at improvement for product types now:
%% \begin{alignat*}{3}
%%   &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Up A,\,\Up B)\,\where}\\
%%   &\hspace{-4em}\ind \mup\,x   && = (\ql \fst\,\spl x \qr,\, \ql \snd\,\spl x \qr)\\
%%   &\hspace{-4em}\ind \mdown\,(a,\,b) && = \ql(\spl a,\, \spl b)\qr
%% \end{alignat*}

@@

\emph{sums-of-products} (SOP in short) later in Section in \ref{TODO}, so we

>>>>

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}
\newcommand{\vl}{\mathsf{l}}
\newcommand{\vns}{\mathsf{ns}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}}
\newcommand{\Tree}{\msf{Tree}}
\newcommand{\Node}{\msf{Node}}
\newcommand{\Leaf}{\msf{Leaf}}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Applicative}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}
\newcommand{\foldr}{\msf{foldr}}
\newcommand{\foldl}{\msf{foldl}}
\newcommand{\rep}{\msf{rep}}
\newcommand{\concatMap}{\msf{concatMap}}

@@

\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mbf{up}}
\newcommand{\mdown}{\mbf{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{{<\!\!*\!\!>}}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\MonadState}{\msf{MonadState}}
\newcommand{\MonadReader}{\msf{MonadReader}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\Either}{\msf{Either}}
\newcommand{\Left}{\msf{Left}}
\newcommand{\Right}{\msf{Right}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}

@@

GHC's optimization efforts are respectable, and it has gotten quite good over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation \cite{ghcdocs}, but without strong
guarantees, and their advanced usage requires knowledge of GHC internals. For
example, correctly specifying the \emph{ordering} of certain rule applications
is often needed. We also have to care about formal function arities. Infamously,
the function composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to
\vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It remains a
common practice in high-performance Haskell programming to manually review GHC's
optimized code output.

@@

\textbf{Two-level type theory} (2LTT) \cite{twolevel,staged2ltt} makes it
possible to use expressive dependent type theories for metaprogramming
for a wide range of object languages. In this paper, we use 2LTT to sweeten the
deal of staged compilation, aiming for a combination of strong guarantees, good
ergonomics, high level of abstraction and easy-to-optimize code output.

We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists
of:
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      in the downstream pipeline, but it lacks many convenience features.
\item A dependent type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}
Since the object language is first-order, we guarantee that all programs in CFTT
can be ultimately compiled without any dynamic closures, using only calls and
jumps to statically known code. Why emphasize closures?  They are the
foundation to almost all abstraction tools in functional programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
\item Functors and first-class modules in OCaml and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

@@

%% \todo{Essential closures: CPS? Threaded interpreters?}
%% Whole-program defunctionalization is also possible, and it is notably used by
%% the MLton compiler \cite{TODO}. While this can be practical and effective, it
%% can be also expensive and require whole-program processing. Also, conceptually
%% speaking, it does not eliminate closures but instead makes them more transparent
%% to optimizations. In this paper we do not use defunctionalization.

@@

\begin{itemize}
\item In Section \ref{sec:overview-of-cftt} we present the two-level type theory
  CFTT, where the object level is first-order simply-typed and the meta level is
  dependently typed. The object language supports an operational semantics without
  runtime closures, and can be compiled in a way such that all function calls
  are statically known. We provide a supplementary Agda formalization of the
  operational semantics.
\item In Section \ref{sec:monad-transformers} we build a monad transformer library. We believe
  that this is a good demonstration, because monads and monad transformers are
  the most widely used effect system in Haskell, and at the same time their
  compilation to efficient code can be surprisingly difficult. The main idea is
  to have monads and monad transformers \emph{only} in the metalanguage, and try
  to express as much as possible at the meta level, and funnel the object-meta
  interactions through specific \emph{binding-time improvements}.  We show how
  almost all definitions from the Haskell ecosystem of monad transformers can be
  reused in the meta level of CFTT. We also develop let-insertion, join points
  and case switching on object-level data, in the generality of monad transformers.
\item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
  we demonstrate essential usage of dependent types, for providing guaranteed
  fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
  state machine representation that is based on \emph{sums-of-products} of
  object-level values. We show that CFTT is compatible with a
  \emph{generativity} axiom, which internalizes the fact that metaprograms
  cannot inspect the structure of object-level terms. We use this to show that
  the universe of sums-of-products is closed under $\Sigma$-types. This in turn
  enables a very concise definition of $\concatMap$.
\item
  We adapt the contents of the paper to typed Template Haskell
  \cite{DBLP:journals/pacmpl/XiePLWYW22}, with some modifications,
  simplifications and fewer guarantees about generated code.  In particular,
  Haskell does not have enough dependent types for the simple $\concatMap$
  definition, but we can still work around this limitation. We also provide a
  precise Agda embedding of CFTT and our libraries as described in this
  paper. Here, the object theory is embedded as a collection of postulated
  operations, and we can use Agda's normalization command to print out generated
  object code.
\end{itemize}

@@

$\bs{\MTy}$ is the universe of types in the compile-time language. We will often
use the term ``metatype'' to refer to inhabitants of $\MTy$, and use
``metaprogram'' for inhabitants of metatypes. $\MTy$ supports dependent
functions, products and indexed inductive types \cite{inductivefamilies}.

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. This is orthogonal to the levels of staging. However,
universe levels add noise and they are not too relevant to the current paper, so
we will omit them.

@@

We use Haskell-like newtype notation, such as in $\newtype \msf{Wrap}\,\vA =
\msf{Wrap}\,\{\mit{unWrap} : \vA \}$, and also a similar notation for (dependent)
record types, for instance as in
\[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]
All construction and elimination rules for type formers in $\MTy$ stay within
$\MTy$. For example, induction on meta-level values can only produce meta-level
values.

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so so we have $\Ty : \MTy$. All construction and elimination rules of
type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
sub-universes.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
to define computations and values alike, and can be used to shadow binders:
\begin{alignat*}{3}
  &\vf : \Nat \to \Nat\\
  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
\end{alignat*}
We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
assumed to be erased at runtime. In the Haskell implementation they are
important for guiding type class resolution, and we think that the explicit
wrapping makes many definitions more comprehensible in CFTT as well.

@@

However, the call-by-name semantics lets us transform $\vf$ to $\lam
\vb\,\vx.\,\case \vb\,\of \True \to \vx + 10; \False \to \vx * 10$, and more
generally we can transform programs so that every function call becomes
\emph{saturated}. This means that every function call is of the form
$\vf\,\vt_1\,\vt_2\,...\,\vt_n$, where $\vf$ is a function variable and the
definition of $\vf$ immediately $\lambda$-binds $n$ arguments. We do not detail
this here. We provide formal syntax and operational semantics of the object
language in the Agda supplement. We formalized the specific translation steps
that are involved in call saturation, but only specified the full translation
informally.

@@

\subsubsection{Comparison to call-by-push-value}
We took inspiration from call-by-push-value (CBPV) \cite{DBLP:conf/tlca/Levy99},
and there are similarities, but there are also significant differences. Both
systems have a value-computation distinction, with call-by-name computations and
call-by-value values. However, our object theory supports variable binding at
arbitrary types while CBPV only supports value variables. In CBPV, a
let-definition for a function is only possible by first packing it up as a
closure value (or ``thunk''), which clearly does not suit our applications.
Investigating the relation between CBPV and our object language could be future work.

%% \subsubsection{Spectrum of possible object languages}
%% There is a trade-off between making the object language more restricted, and
%% thus easier to compile, and making metaprogramming more convenient. We will see
%% that the ability to insert $\mbf{let}$-s without restriction is convenient, and
%% likewise the ability to have $\lambda$-abstraction under $\mbf{case}$ bodies,
%% although these features necessitate more downstream processing, to bring
%% programs to saturated form. On the other side of the spectrum, one might imagine
%% generating object code in low-level A-normal form. This is more laborious, but
%% it could be also interesting, because it forces us to invent abstractions for
%% manipulating ANF in the metalanguage. In a similar vein, Allais recently
%% proposed metaprogramming quantum circuits in a two-level type theory
%% \cite{TODO}.

@@

\begin{itemize}
\item For $\vA : \Ty$, we have $\Lift \vA : \MTy$, pronounced as ``lift $\vA$''. This is
      the type of metaprograms that produce $\vA$-typed object programs.
\item For $\vA : \Ty$ and $\vt : \vA$, we have $\ql \vt \qr : \Lift \vA$, pronounced ``quote $\vt$''. This
      is the metaprogram which immediately returns $\vt$.
\item For $\vt : \Lift \vA$, we have $\spl \vt : A$, pronounced ``splice
  $\vt$''. This inserts the result of a metaprogram into an object
  term. \emph{Notation:} splicing binds stronger than function application, so
  $\vf\,\spl \vx$ is parsed as $\vf\,(\spl \vx)$. We borrow this notation from
  MetaML \cite{metaml}.
\item We have $\ql \spl \vt \qr \equiv \vt$ and $\spl \ql \vt \qr \equiv \vt$ as definitional equalities.
\end{itemize}

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. \textbf{Staging} means running all metaprograms in splices and
inserting their output into object code, keeping all object-level top entries
and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
and produces output which is purely in the object-level fragment, with no
metatypes and metaprograms remaining. The output is guaranteed to be well-typed.

@@

\subsection{Semantics of Staging}\label{sec:semantics-of-staging}

We give a short summary of the semantics of staging here. We reuse most of the
semantics from \cite{staged2ltt} and only mention additions and differences. The
reader may refer to ibid.\ for technical background.

Staging is given by evaluation of CFTT into a presheaf model, where the base
category is given as follows: objects are typing contexts of the object language
and morphisms are parallel substitutions. The equality of morphisms is given by
strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
which is constantly the set of object-theoretic types. Soundness of staging is
proven by a logical relation model, internally to the mentioned presheaf
model. Soundness means that the output of staging, viewed as a CFTT term, is
definitionally equal to the input.

The main addition in CFTT is the assumption of all inductive families in the
meta-level. However, based on the results in \cite{whynotw}, it is enough to
extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
derive all inductive families. We consider the semantics of these in the following.

\subsubsection{W-types} W-types have a standard object-wise inductive definition in
the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
interpretation to W-types, as required for the soundness proof for staging. In
Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
for natural numbers, however, there is no analogous map for W-types because we
cannot build finite terms from infinitely branching trees.

@@

\subsubsection{Identity type} Our treatment of the identity type is
influenced by the following: we would like to assume equations as axioms in
CFTT, which are true in the presheaf model (i.e.\ they are true at
staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
we use such an axiom to good effect. However, axioms do block computation in
CFTT up to definitional equality, which implies that the soundness of staging
fails. Our solution:
\begin{itemize}
\item Have a standard intensional identity type in CFTT.
\item Before staging, translate the intensional identity to extensional
      identity. This amounts to erasing all identity proofs and transports.
\item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
\end{itemize}
In the presheaf model, we interpret the extensional identity in a standard way,
as strict equality of sections of presheaves, and soundness for extensional
identity becomes trivial to show. Hence, the soundness property for our overall
staging pipeline becomes the following: the staging output is definitionally
equal to the identity-erased version of the staging input.

@@

In this section we build a library for monads and monad transformers. We believe
that this is a good demonstration of CFTT-s abilities, since monads are
ubiquitous in Haskell programming and they also introduce a great amount of
abstraction that should be optimized away.

@@

In the staged compilation and partial evaluation literature, the term
\emph{binding time improvement} is used to refer to such conversions, where the
``improved'' version supports more compile-time computation
\cite{partial-evaluation}. A general strategy for generating efficient ``fused''
programs, is to try to work as much as possible with improved representations,
and only convert back to object code at points where runtime dependencies are
unavoidable. Let us look at binding-time-improvement for product types now:
\begin{alignat*}{3}
  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
\end{alignat*}

@@

\emph{sums-of-products} ($\SOP$ in short) later in Section in \ref{sec:stream-fusion}, so we
